Data migration under a Changing Schema

### 2023-01-08 2022 Migration/articleMigration.tex
changed sct 3.7 Loading data

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}

@@

   \item The generator generates a migration script $\migrsys$

@@

         A migration engineer changes $\migrsys$ to accommodate the non-standard requirements of the migration,
         yielding migration script $\migrsys'$.
         The migration engineer does not alter the existing system $\infsys$ nor the specification
         of the desired system $\infsys'$ that the developer has produced.

@@

   The relabeling of concepts is not trivial because of specialization.
   To illustrate this point, suppose $D\isa A$.
   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
   For example by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
   all atoms that represent hotels might suddenly be expected to have parking lots.
   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
\subsection{Example}
   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
   We will use the example in section~\ref{sct:Example existing IS} as the existing system
   and call that $\infsys$.
   Its population reflects the state of the system just before the migration.

   Information system $\infsys'$, the desired system, is specified by:
\begin{verbatim}
   RELATION takes[Student*Course]
   RELATION isPartOf[Module*Course]
   RELATION isEnrolledFor[Student*Module]
   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
   ROLE Administrator MAINTAINS EnrollRule

   RELATION course[ExamReg*Course] [UNI] =
      [ ("ER1", "Management")
      ; ("ER2", "Business IT")
      ; ("ER3", "Business IT")
      ]
   RELATION student[ExamReg*Student] [UNI] =
      [ ("ER1", "Peter")
      ; ("ER2", "Susan")
      ]
   RULE ExamRule1: student~;course |- takes
   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
\end{verbatim}
   We can see that the relations and the rule of the existing system remain part of the desired system%
\footnote{Removal or changing of relations is allowed as well.}
   because the developer has copied these items into the script of the desired system.
   However, she has lifted the restriction \verb-UNI- on the relation {\tt isPartOf}.
   Besides, two new relations and two rules have appeared.

   The desired system, $\infsys'$, contains exam registrations (concept \verb-ExamReg-), which is new.
   In an exam registration, a student registers for the examination of a course.
   $\infsys'$ contains two extra rules: \verb-ExamRule1- and \verb-ExamRule2-.
   The first one says that an exam registration requires that a student actually takes the course.
   In logic, the constraint $p_{\tt ExamRule1}$ is written as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
\end{array}\]
   Another requirement, \verb-ExamRule2-, is that the student is enrolled for every module that is part of the course.
   Its constraint, $p_{\tt ExamRule2}$, is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
\end{array}\]
   % Ampersand also derives the following violation sets:
% \[\begin{array}{l}
   % \viol{\text{\tt ExamRule1}}{\dataset}\\
   % \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
   % \viol{\text{\tt ExamRule2}}{\dataset}\\
   % \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
% \end{array}\]

\subsection{Copying Data}
   For loading the data in the desired database, we use enforce rules.
   In Ampersand, an enforce rule comes in one of the following three kinds:
\begin{eqnarray}
   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{expression}\label{enforce del}\\
   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{expression}\label{enforce eq}\\
   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{expression}\label{enforce ins}
\end{eqnarray}
   Every enforce rule is enforced by an engine,
   which changes the population of relation $r$ to satisfy the rule.
   An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
   When the expression is losing pairs that are in $r$, the engine will spontaneously remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
   An enforce rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
   When the expression is growing with pairs that are in $r$, the engine will spontaneously insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression}.
   It does so by inserting and deleting pairs in $r$ to keep it equal to \id{expression}.
   Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.

   It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
The prefix {\tt new.} refers to items from the desired information system (this section).}:
\begin{verbatim}
   ENFORCE new.takes >: old.takes
\end{verbatim}
   This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
   An extra relation fixes this problem by remembering which pairs have been copied:

\begin{verbatim}
   RELATION copied_takes[Student*Course]
   ENFORCE new.takes >: old.takes - copied_takes
   ENFORCE copied_takes >: new.takes
\end{verbatim}
   Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
   So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes}.
   Then, the second enforce rule will start filling {\tt copied\_takes} with pairs from {\tt new.takes},
   and the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
   Then, the enforce rule that fills {\tt new.takes} will be satisfied forever and will change nothing more to {\tt new.takes}.

   So, the generator that ``writes'' the code for the migration system must introduce an extra relation for each relation to be copied.
   After the migration has finished, this extra relation and both enforce rules can be taken out of production.
   For this purpose, the generator writes a rule that signals the migration engineer when copying is done.

\subsection{Migration system}
   The generated migration system is:
\begin{verbatim}
   RELATION copied_takes[Student*Course]
   ENFORCE new.takes >: old.takes - copied_takes
   ENFORCE copied_takes >: new.takes

@@

\subsection{Moment of Transition}
   Let us assume the population just before the moment of transition is.
\begin{verbatim}
   RELATION takes[Student*Course]
   RELATION isPartOf[Module*Course] [UNI]=
      [ ("IT-Governance", "Business IT") ]
   RELATION isEnrolledFor [Student*Module] =
      [ ("Susan", "Business Analytics")
      ; ("Susan", "IT-Governance")
      ; ("Susan", "Business Rules")
      ]
\end{verbatim}

>>>>

   The example system also has a second relation that states which modules are part of which course.
\begin{verbatim}
RELATION isPartOf[Module*Course] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}

@@

   \item The generator generates a migration script, $\migrsys$,

@@

         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.
         However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
         For that purpose, the generator generates the migration script as an Ampersand script,
         to let the migration engineer accommodate $\migrsys$ to the non-standard requirements.
         The resulting migration script, $\migrsys'$, can be tested separately and can be used to take the desired system to production.
         The migration engineer does not alter the schemas of the existing system or the desired system.
         All migration functionality is concentrated in $\migrsys'$.

@@

   The relabeling of concepts is not trivial because of specialization.
   To illustrate this point, suppose $D\isa A$.
   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
   For example by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
   all atoms that represent hotels might suddenly be expected to have parking lots.
   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.

\subsection{Loading Data}
   For loading the data in the desired database, we use enforce rules.
   In Ampersand, an enforce rule comes in three flavors:
\begin{eqnarray}
   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{expression}\label{enforce del}\\
   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{expression}\label{enforce eq}\\
   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{expression}\label{enforce ins}
\end{eqnarray}
   Every enforce rule is enforced by an engine,
   which changes the population of relation $r$ to satisfy the rule.
   An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
   When the population of this expression is losing pairs that are in $r$, the engine will remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
   An enforce rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
   When the population of this expression is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression}.
   It does so by inserting and deleting pairs in $r$ to keep it equal to \id{expression}.
   Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.

   It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
The prefix {\tt new.} refers to items from the desired information system.}:
\begin{verbatim}
   ENFORCE new.takes >: old.takes
\end{verbatim}
   This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
   An extra relation (\ref{extraRelation}) fixes this problem by remembering which pairs have been copied:
\begin{eqnarray}
   &&\verb#RELATION copied_takes[Student*Course]#\label{extraRelation}\\
   &&\verb#ENFORCE new.takes >: old.takes - copied_takes#\label{difference}\\
   &&\verb#ENFORCE copied_takes := new.takes#\label{fill copies with new.takes}\\
   &&\verb#ENFORCE copied_takes :< old.takes#\label{del copies from old.takes}
\end{eqnarray}
   Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
   So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes} by rule~\ref*{difference}.
   Then, the engine will start filling {\tt copied\_takes} with pairs from {\tt new.takes} because of rule~\ref*{fill copies with new.takes}.
   As a result, the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
   While this copying process is going on, {\tt old.takes} may accumulate more pairs or lose pairs.
\begin{itemize}
\item If pair $p$ is inserted in {\tt old.takes} and $p$ is not in {\tt copied\_takes},
      then rule~\ref{difference} causes the engine to insert $p$ in {\tt new.takes}.
      Subsequently, rule~\ref{fill copies with new.takes} causes the engine to insert $p$ in {\tt copied\_takes} as well.
\item If pair $p$ is inserted in {\tt old.takes} and $p$ is already in {\tt copied\_takes},
      then nothing happens.
\item If pair $p$ is deleted from {\tt old.takes} and $p$ is in {\tt copied\_takes},
      then rule~\ref{del copies from old.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
\item If pair $p$ is deleted from {\tt old.takes} and $p$ is not in {\tt copied\_takes},
      then nothing happens.
\item If pair $p$ is inserted in {\tt new.takes},
      rule~\ref{fill copies with new.takes} causes the engine to insert $p$ into {\tt copied\_takes}.
\item If pair $p$ is deleted from {\tt new.takes},
      rule~\ref{fill copies with new.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
      If this pair is in {\tt old.takes}, rule~\ref{difference} causes the engine to insert $p$ back into {\tt new.takes}.
\end{itemize}

   These new pairs are copied by the same rule.
   However, {\tt old.takes} may also lose pairs, which are 
   Then, the enforce rule that fills {\tt new.takes} will be satisfied forever and will change nothing more to {\tt new.takes}.

   But what if (in the meantime) someone removes a pair from {\tt old.takes} that has already been inserted in {\tt new.takes}?
\begin{verbatim}
   RELATION added_takes[Student*Course]
   RELATION deleted_takes[Student*Course]
   ENFORCE new.takes >: old.takes - added_takes
   ENFORCE added_takes >: new.takes
\end{verbatim}
   So, the generator that ``writes'' the code for the migration system must introduce an extra relation for each relation to be copied.
   After the migration has finished, this extra relation and both enforce rules can be taken out of production.

\subsection{Example}
   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
   We will use the example in section~\ref{sct:Example existing IS} as the existing system
   and call that $\infsys$.
   Its population reflects the state of the system just before the migration.

   Information system $\infsys'$, the desired system, is specified by:
\begin{verbatim}
   RELATION takes[Student*Course]
   RELATION isPartOf[Module*Course] [UNI]
   RELATION isEnrolledFor[Student*Module]
   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
   ROLE Administrator MAINTAINS EnrollRule

   RELATION course[ExamReg*Course] [UNI] =
      [ ("ER1", "Management")
      ; ("ER2", "Business IT")
      ; ("ER3", "Business IT")
      ]
   RELATION student[ExamReg*Student] [UNI] =
      [ ("ER1", "Peter")
      ; ("ER2", "Susan")
      ]
   ROLE Student MAINTAINS ExamRule1, ExamRule2
   RULE ExamRule1: student~;course |- takes
   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
\end{verbatim}
   We can see that the relations and the rule of the existing system remain part of the desired system%
\footnote{Removal or changing of relations is allowed as well.}
   because the developer has copied these items into the script of the desired system.
   However, she has imposed the restriction \verb-UNI- on the relation {\tt isPartOf}, stating that this relation is univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
   Besides, two new relations and two rules have appeared.

   The desired system, $\infsys'$, contains exam registrations (concept \verb-ExamReg-), which is new.
   In an exam registration, a student registers for the examination of a course.
   $\infsys'$ contains two extra rules: \verb-ExamRule1- and \verb-ExamRule2-.
   The first one says that an exam registration requires that a student actually takes the course.
   In logic, the constraint $p_{\tt ExamRule1}$ is written as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
\end{array}\]
   Another requirement, \verb-ExamRule2-, is that the student is enrolled for every module that is part of the course.
   Its constraint, $p_{\tt ExamRule2}$, is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
\end{array}\]
   Both rules must be maintained by the student.
   So, a student can register for an exam only if (s)he is taking the course for which (s)he registers and if (s)he is enrolled for every module of that course.
   % Ampersand also derives the following violation sets:
% \[\begin{array}{l}
   % \viol{\text{\tt ExamRule1}}{\dataset}\\
   % \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
   % \viol{\text{\tt ExamRule2}}{\dataset}\\
   % \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
% \end{array}\]

\subsection{Migration system}
   The generated migration system, $\migrsys$, is:
\begin{verbatim}
   RELATION copied_takes[Student*Course]
   ENFORCE new.takes >: old.takes - copied_takes
   ENFORCE copied_takes >: new.takes

@@

\subsection{Preparing for the Moment of Transition}
   Just before loading the desired system with data,
   the data in the existing system is no longer the data we saw in section~\ref{sct:Example existing IS}.
   Let us assume the actual population at this time is:
\begin{verbatim}
   RELATION takes[Student*Course] =
      [ ("Peter", "Management")
      ; ("Susan", "Business IT")
      ; ("John", "Business IT")
      ]
   RELATION isPartOf[Module*Course] [UNI]=
      [ ("Finance", "Management")
      ; ("Business Rules", "Business IT")
      ; ("Business Analytics", "Business IT")
      ; ("IT-Governance", "Management")
      ; ("IT-Governance", "Business IT")
      ]
   RELATION isEnrolledFor [Student*Module] =
      [ ("Susan", "Business Analytics")
      ; ("Susan", "IT-Governance")
      ; ("Susan", "Business Rules")
      ]
\end{verbatim}
